\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-11-17 review_response.Rnw

*Main text*

I didn't find it obvious from the text whether the setstep and comb functions would handle missing data with interpolation or NA. (It's NA, as it turns out.) Perhaps it is because I don't do a lot of work in this domain, though.

{\it This was clarified in the revision, page 6 paragraph one:

`Both functions treat missing data as NA values, either for observations that exceed the allowable tolerance for the differ argument of setstep or for portions of time series that do not overlap given the method argument passed to comb.' 
}

@@

*Package performance*

Some of the functions can spend a surprising amount of time parsing. For example, the call: all\_params\_dtrng(`hudscwq', dtrng = c(`09/01/2013', `10/01/2013')) took 19 seconds on my machine, and running under a profiler showed that about 12-3 of those seconds were spent parsing the data, not downloading it. The SWMPr::parser code calls htmlTreeParse, which is where all that time is being spent; I found that changing it to xmlTreeParse made almost all of that time go away. The only difference I noticed was that htmlTreeParse converts all of the tag names to lowercase, while xmlTreeParse does not. This can be easily fixed by having the parser() function change the `out' data frame's names to lowercase before returning. With these two modifications, the elapsed time drops to 6-7 seconds--the bulk of the time now being spent downloading the data (as far as I can tell).

{\it That is a fantastic suggestion and I've made these changes in the development repository of the package (see \href{https://github.com/fawda123/SWMPr/commit/3a90bddabbeace917b54942a68449ef9be93ad1d}{commit}).  I noticed an immediate increase in performance with the import functions.  I will push the new version to CRAN if or when the paper is accepted for publication. 
}

@@

{\it Again, these are all great suggestions but I would probably need assistance to implement these changes with the new API.}

*shiny\_summary app*

In the file https://github.com/fawda123/swmp\_summary/blob/master/server.R, the plotInput should not be a reactive(), but a regular function that takes no arguments. This is because reactives cache their values, and should not be executed for their side effects (in this case, plotting). As the app is written now, there doesn't appear to be an actual user-visible bug, but a change as simple as setting the plot width to ``100\%'' would introduce a bug (I'd expect it to fail to redraw the plot when the window was resized).

{\it Yes, the plot fails to render on shinyapps when the window is resized.  I had not noticed the bug so I appreciate the comment.  I've changed the code in server.R to make plotInput a regular function (see \href{https://github.com/fawda123/swmp_summary/commit/907abdd25b543cd5db372a8f30ecd9d7e49611e7}{commit}).  I've redeployed the app and the bug has been fixed.}

@@

\Bigtxt{Reviewer 2:}

First, I did not see any information about how SWMPR treats censored data, i.e., measurements that can be reported only as less than or greater than an analysis or reporting limit. For example, measurements of nutrient parameters could include values that are below analytical detection limits. A considerable amount of research has shown that the assumptions and methods used to analyze data that include censored values can influence the conclusions of the analyses (for example, see Helsel 2012). Considerable attention is given in the manuscript, appropriately, to various quality assurance/quality control functions and flags, and I realize not every aspect of QA/QC needs to be covered explicitly in the manuscript. However, I believe that the topic of censored values is important enough that the paper would be improved by providing explicit information about how censored values may be coded, retrieved, and ultimately used in various data analyses carried out by SWMPr.

{\it I fully agree that considering censored data is a necessary aspect of processing and analyzing water quality data.  Censored data are given unique QAQC flags during initial data processing through the Centralized Data Management Office (CDMO).  Censored values in the imported data can be identified using functions in the SWMPr package, although the previous draft did not sufficiently describe this information.  I believe that the treatment of censored values is beyond the scope of the SWMPr package as there are already methods available in R for working with these observations.  However, I agree that the issue should be stated more clearly in the manuscript and, more importantly, how SWMPr can be used to identy censored data.  I have revised the text accordingly (page 4, paragraph 3):

``SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.  For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.  The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series (Helsel, 2012).  A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages (e.g., cents, McLeod et al., 2014).  However, the functions in SWMPr can be used to identify censored data based on the appropriate QAQC flag for a parameter.  Viewing this information can be helpful for determining how to further process the data with the qaqc function or alternative methods outside of SWMPr.  The qaqcchk function returns a data.frame of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.  SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.''
}

@@

Second, in the ``Applications using the SWMPr package'' section near the end of the manuscript, simple linear regression (SLR) is identified as a technique used to summarize trends in a parameter over time. This particular statistical method may or may not be appropriate for a given data set. Alternative trend analysis methods exist, including several that treat censored values appropriately (again, Helsel 2012 is a useful reference). It is unclear from the manuscript what alternatives to SLR may be available in SWMPr (e.g., maximum likelihood estimation, Akritas-Theil-Sen nonparametric regression, others). Therefore, clarifying text that either identifies the availability of other trend analysis procedures in SWMPr or provides caveats to the use of SLR (e.g., used only for initial exploratory analysis) should be added to this section.

{\it I agree that the use of SLR is a simple method that may not be appropriate for all datasets provided by SWMP.  The application is meant to be exploratory and I have revised the text to clarify that SLR may be an inadequate method for trend analysis, particularly with censored data.  The text description on the application was also revised.

Page 12, paragraph one: `More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.'

### 2015-11-17 swmpr_manu.Rnw
response to first reviewer done, need to proofread, recompile
edits before resubmit

\abstract{
The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. However, the increasing quantity of data provided by the monitoring network have complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.  This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
}

@@

SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.  For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.  The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for the time series \citep{Helsel12}.  A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.  However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a given parameter.  Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.  The \code{qaqcchk} function returns a \code{data.frame} of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.  SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.

@@

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.  A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.  As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.

@@

# combine nut and wq data by intersect
comb(swmp1, swmp3, method = 'intersect')

@@

descrips <- c(
  'Aggregate \\code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
  'Aggregate metabolism data from a \\code{"swmpr"} object.  This is primarly used within \\code{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \\citep{Odum56}.',
  'Decompose a \\code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\code{decompose} \\citep{Kendall83}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\code{decompTs} in the \\CRANpkg{wq} package \\citep{Jassby14}.  Only monthly decomposition is possible.',
  'Plot a histogram for a single variable.',
  'Add lines to an existing plot created with \\code{plot}.',
  'Create a map of all stations in a reserve using the \\CRANpkg{ggmap} package \\citep{Kahle13}.',
  'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object.', 
  'Plot multiple time series on the same y-axis.', 
  'Plot a univariate  time series for a \\code{"swmpr"} object.',
  'Plot ecosystem metabolism estimates after running \\code{ecometab} on a combined \\code{"swmpr"} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a single parameter.',
  'Smooth \\code{"swmpr"} objects with a moving window average, passed to \\code{filter}.'
  )

@@

Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.  This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis.

@@

Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}. The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.

@@

Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.  The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.  More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for exploratory purposes only.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}

@@

I acknowledge the significant work of NERRS researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions. Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.

>>>>

\abstract{
The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. However, the increasing quantity of data provided by the monitoring network has complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.  This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
}

@@

SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.  For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.  The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series \citep{Helsel12}.  A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.  However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a parameter.  Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.  The \code{qaqcchk} function returns a \code{data.frame} of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.  SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.

@@

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  Data are combined by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{'union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects, whereas \code{'intersect'} will create a continuous time series from the set of dates that are shared between input objects.  A character string or numeric vector can also be used to specify which of the input objects to use as the master time series for combining.  As with \code{setstep}, a default method for \code{comb} is provided to allow use with arbitrary data structures.  Both functions treat missing data as \code{NA} values, either for observations that exceed the allowable tolerance for the \code{differ} argument of \code{setstep} or for portions of time series that do not overlap given the \code{method} argument passed to \code{comb}.

@@

# combine nut and met data by intersect
comb(swmp1, swmp3, method = 'intersect')

@@

descrips <- c(
  'Aggregate \\code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
  'Aggregate metabolism data from a \\code{"swmpr"} object.  This is primarly used within \\code{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \\citep{Odum56}.',
  'Decompose a \\code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\code{decompose} \\citep{Kendall83}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\code{decompTs} in the \\CRANpkg{wq} package \\citep{Jassby14}.  Only monthly decomposition is possible.',
  'Plot a histogram for a single variable.',
  'Add lines to an existing plot created with \\code{plot}.',
  'Create a map of all stations in a reserve using the \\CRANpkg{ggmap} package \\citep{Kahle13}.',
  'Linearly interpolate missing data (\\code{NA} values) in a \\code{"swmpr"} object.', 
  'Plot multiple time series in a \\code{"swmpr"} object on the same y-axis.', 
  'Plot a univariate  time series for a \\code{"swmpr"} object.',
  'Plot ecosystem metabolism estimates after running \\code{ecometab} on a combined \\code{"swmpr"} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a single parameter.',
  'Smooth \\code{"swmpr"} objects with a moving window average, passed to \\code{filter}.'
  )

@@

Similarly, the \code{overplot} function provides an alternative approach to viewing observed data from the same station.  This function uses the base \pkg{graphics} package to plot multiple time series on the same y-axis (Figure~\ref{fig:overplot}).

@@

Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package (Figure~\ref{fig:map_ex}, \citet{Kahle13}). The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.

@@

Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.  The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.  More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.

\begin{figure}[!h]
\begin{center}
\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}

@@

I acknowledge the significant work of NERRS researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions. Thanks to Jeff Hollister for providing useful comments on an earlier draft.
